---
title: "DATA 607 — Week 2A: SQL and R (Movie Ratings)"
author: "Nawel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DBI)
library(RSQLite)
```

## Overview

This report collects movie ratings for six titles from six people, stores them
in a small relational SQL database (`users`, `movies`, `ratings`), and loads
that data into R for analysis. The goal is to practice the full round trip —
SQL schema design, a many-to-many ratings table, and reading SQL output into a
tidy R data frame — while handling the reality that not everyone has seen
every movie. Full SQL source:
[create_tables.sql](https://raw.githubusercontent.com/amine-souissi0/data607-week2/main/week2a_movie_ratings/create_tables.sql).

> **Note on the data:** the ratings below are placeholder values while the
> real 5-person survey is being collected. Only `create_tables.sql` needs to
> change to swap in real responses — this report re-runs unchanged.

## Loading the SQL Script from GitHub

The database is *built at run time* from a `.sql` file hosted on GitHub, so
the whole pipeline is reproducible from a URL rather than a local file: R
downloads the raw SQL text, then executes each statement against a fresh,
in-memory SQLite database.

```{r load-sql}
sql_url <- "https://raw.githubusercontent.com/amine-souissi0/data607-week2/main/week2a_movie_ratings/create_tables.sql"

sql_lines <- readLines(sql_url, warn = FALSE)

# Drop full-line comments BEFORE splitting into statements. Doing this after
# splitting would fail: this file opens with a multi-line comment block, so
# the first "chunk" between semicolons starts with a comment line even
# though it ends with a real CREATE TABLE statement — checking only whether
# the whole chunk starts with "--" would wrongly discard it.
sql_text <- sql_lines |>
  discard(~ str_starts(str_trim(.x), "--")) |>
  paste(collapse = "\n")

statements <- str_split(sql_text, ";")[[1]] |>
  str_trim() |>
  discard(~ .x == "")

con <- dbConnect(RSQLite::SQLite(), ":memory:")
walk(statements, ~ dbExecute(con, .x))
```

## Querying the Database into a Data Frame

A `JOIN` across all three tables turns the normalized SQL tables back into one
wide, human-readable table — this is the "user-item matrix" the assignment
mentions.

```{r query}
ratings_df <- dbGetQuery(con, "
  SELECT u.name  AS user_name,
         m.title AS movie_title,
         r.rating
  FROM ratings r
  JOIN users  u ON u.user_id  = r.user_id
  JOIN movies m ON m.movie_id = r.movie_id
")

ratings_df <- ratings_df |> as_tibble()
head(ratings_df, 10)
```

## Handling Missing Ratings

Not every person has seen every movie, so `rating` can be `NA`. We keep those
rows (they're informative — "hasn't seen it" is different from "hated it")
but exclude them from any average.

```{r missing-data}
ratings_df |>
  summarize(
    total_pairs   = n(),
    missing_count = sum(is.na(rating)),
    missing_pct   = round(100 * mean(is.na(rating)), 1)
  )
```

## Analysis

**Average rating per movie** (missing ratings excluded via `na.rm = TRUE`):

```{r avg-per-movie}
ratings_df |>
  group_by(movie_title) |>
  summarize(
    avg_rating = round(mean(rating, na.rm = TRUE), 2),
    n_ratings  = sum(!is.na(rating))
  ) |>
  arrange(desc(avg_rating))
```

**Average rating given per user** — a quick check for whether someone rates
systematically higher or lower than the group:

```{r avg-per-user}
ratings_df |>
  group_by(user_name) |>
  summarize(
    avg_given = round(mean(rating, na.rm = TRUE), 2),
    n_seen    = sum(!is.na(rating))
  ) |>
  arrange(desc(avg_given))
```

```{r cleanup, include=FALSE}
dbDisconnect(con)
```

## Conclusions and Findings

With the current placeholder data, ratings cluster in the 3–5 range with no
extreme outliers, and missing ratings are spread across both users and movies
rather than concentrated in one of either — a sign the "hasn't seen it" gaps
are close to random rather than a sign of a bad question or a bad movie
choice. To extend this work: (1) replace the placeholder `INSERT`s with real
survey responses, (2) add a `genre` column to `movies` to check whether
average rating varies by genre, and (3) reshape `ratings_df` into a wide
user-by-movie matrix (`pivot_wider`) as a first step toward a recommender
system, since that matrix is exactly what collaborative filtering needs as
input.

## AI Use

Anthropic. (2026). Claude (model: claude-sonnet-5) [Large language model].
[https://claude.ai/](https://claude.ai/) — used to design the SQL schema
(users/movies/ratings junction table), write the R code that loads the SQL
script from a URL and executes it against an in-memory SQLite database, and
draft the analysis and this write-up. Reviewed and adapted by the author.
