---
title: "DATA 607 — Week 2A: SQL and R (Movie Ratings)"
author: "Nawel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(DBI)
library(RSQLite)
```

## Overview

This report collects ratings for six titles across five raters, stores them
in a small relational SQL database (`users`, `movies`, `ratings`), and loads
that data into R for analysis. The goal is to practice the full round trip —
SQL schema design, a many-to-many ratings table, and reading SQL output into a
tidy R data frame — while handling the reality that not every rater has an
opinion on every title. Full SQL source:
[create_tables.sql](https://raw.githubusercontent.com/amine-souissi0/data607-week2/main/week2a_movie_ratings/create_tables.sql).

> **Data source disclosure:** for this assignment I used a small
> representative sample derived from publicly available aggregate ratings on
> Goodreads and IMDb, rather than a live survey of five people I personally
> asked. IMDb reports *Grey's Anatomy* at 7.6/10 and *Titanic* at 8.0/10,
> which convert to approximately 3.8/5 and 4.0/5; Goodreads reports *Perfume*
> at 4.04/5 and *The Little Prince* at 4.34/5. The five rows in `ratings`
> were constructed so each title's average matches its public score, with a
> couple of cells left missing (`NULL`) to model a rater who hasn't seen or
> read that title. The rows do not represent people I personally surveyed;
> ratings are on a 1–5 scale, and unavailable ratings are stored as missing
> values rather than as zero.

## Loading the SQL Script from GitHub

The database is *built at run time* from a `.sql` file hosted on GitHub, so
the whole pipeline is reproducible from a URL rather than a local file: R
downloads the raw SQL text, then executes each statement against a fresh,
in-memory SQLite database.

```{r load-sql}
sql_url <- "https://raw.githubusercontent.com/amine-souissi0/data607-week2/main/week2a_movie_ratings/create_tables.sql"

sql_lines <- readLines(sql_url, warn = FALSE)

# Drop full-line comments BEFORE splitting into statements. Doing this after
# splitting would fail: this file opens with a multi-line comment block, so
# the first "chunk" between semicolons starts with a comment line even
# though it ends with a real CREATE TABLE statement — checking only whether
# the whole chunk starts with "--" would wrongly discard it.
sql_text <- sql_lines |>
  discard(~ str_starts(str_trim(.x), "--")) |>
  paste(collapse = "\n")

statements <- str_split(sql_text, ";")[[1]] |>
  str_trim() |>
  discard(~ .x == "")

con <- dbConnect(RSQLite::SQLite(), ":memory:")
walk(statements, ~ dbExecute(con, .x))
```

## Querying the Database into a Data Frame

A `JOIN` across all three tables turns the normalized SQL tables back into one
wide, human-readable table — this is the "user-item matrix" the assignment
mentions.

```{r query}
ratings_df <- dbGetQuery(con, "
  SELECT u.name  AS user_name,
         m.title AS movie_title,
         r.rating
  FROM ratings r
  JOIN users  u ON u.user_id  = r.user_id
  JOIN movies m ON m.movie_id = r.movie_id
")

ratings_df <- ratings_df |> as_tibble()
head(ratings_df, 10)
```

## Handling Missing Ratings

Not every person has seen every movie, so `rating` can be `NA`. We keep those
rows (they're informative — "hasn't seen it" is different from "hated it")
but exclude them from any average.

```{r missing-data}
ratings_df |>
  summarize(
    total_pairs   = n(),
    missing_count = sum(is.na(rating)),
    missing_pct   = round(100 * mean(is.na(rating)), 1)
  )
```

## Analysis

**Average rating per movie** (missing ratings excluded via `na.rm = TRUE`):

```{r avg-per-movie}
ratings_df |>
  group_by(movie_title) |>
  summarize(
    avg_rating = round(mean(rating, na.rm = TRUE), 2),
    n_ratings  = sum(!is.na(rating))
  ) |>
  arrange(desc(avg_rating))
```

**Average rating given per user** — a quick check for whether someone rates
systematically higher or lower than the group:

```{r avg-per-user}
ratings_df |>
  group_by(user_name) |>
  summarize(
    avg_given = round(mean(rating, na.rm = TRUE), 2),
    n_seen    = sum(!is.na(rating))
  ) |>
  arrange(desc(avg_given))
```

```{r cleanup, include=FALSE}
dbDisconnect(con)
```

## Conclusions and Findings

*The Little Prince* comes out on top (4.4) and *Lolita*/*Grey's Anatomy* tie
for lowest (3.8), which simply reproduces the public scores each row was
built from — with only five rows and two missing cells, this dataset is too
small to draw any real conclusion about rater behavior beyond confirming the
pipeline (schema, JOIN, missing-value handling) works correctly end to end.
Because the ratings are a constructed sample rather than five independently
surveyed people, they shouldn't be read as evidence of genuine disagreement
between raters — that's the main limitation to flag if this were extended.
To extend this work: (1) replace these constructed rows with a live survey of
five real people, which would make patterns like "who rates highest on
average" actually meaningful, (2) add a `genre` column to `movies` to check
whether average rating varies by genre, and (3) reshape `ratings_df` into a
wide user-by-movie matrix (`pivot_wider`) as a first step toward a
recommender system, since that matrix is exactly what collaborative
filtering needs as input.

## AI Use

Anthropic. (2026). Claude (model: claude-sonnet-5) [Large language model].
[https://claude.ai/](https://claude.ai/) — used to design the SQL schema
(users/movies/ratings junction table), write the R code that loads the SQL
script from a URL and executes it against an in-memory SQLite database, and
draft the analysis and this write-up. Reviewed and adapted by the author.
