---
title: "DATA 607 — Week 2B: Classification Metrics"
author: "Nawel"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## Overview

This report evaluates a binary classifier that predicts penguin sex (female
vs. male) from the Palmer Penguins dataset, and looks at how the choice of
probability threshold changes what the model's performance metrics say.
`.pred_female` is the model's predicted probability of "female"; `sex` is the
true label. Data:
[penguin_predictions.csv](https://raw.githubusercontent.com/amine-souissi0/data607-week2/main/week2b_classification_metrics/penguin_predictions.csv).

```{r load-data}
data_url <- "https://raw.githubusercontent.com/amine-souissi0/data607-week2/main/week2b_classification_metrics/penguin_predictions.csv"
penguins <- read_csv(data_url, show_col_types = FALSE)
glimpse(penguins)
```

## Null Error Rate

The null error rate is the error you'd get by ignoring the model entirely and
always guessing the majority class. It matters because it's the baseline a
real model has to beat — a model that's 90% accurate sounds good until you
learn the majority-class guess alone gets 85%.

```{r null-error-rate}
class_counts <- penguins |> count(sex)
class_counts

majority_class <- class_counts |> slice_max(n, n = 1) |> pull(sex)
null_error_rate <- 1 - max(class_counts$n) / sum(class_counts$n)

cat("Majority class:", majority_class, "\n")
cat("Null error rate:", round(null_error_rate, 4), "\n")
```

```{r class-distribution-plot}
ggplot(penguins, aes(x = sex, fill = sex)) +
  geom_bar() +
  labs(
    title = "Distribution of actual class (sex)",
    x = "Sex", y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

With 54 male and 39 female penguins (of 93 total), always guessing "male"
would be right 58.1% of the time — a null error rate of 41.9%. Any model we
build must beat 41.9% error to be worth using at all.

## Confusion Matrices at Multiple Thresholds

`.pred_class` in the raw data was already computed at the default 0.5
threshold. To compare thresholds fairly we recompute the predicted class
ourselves from `.pred_female`, rather than trusting the pre-computed column,
so the same logic applies at 0.2 and 0.8 too.

```{r confusion-function}
# Treat "female" as the positive class, since .pred_female is P(female).
confusion_at <- function(df, threshold) {
  df |>
    mutate(pred = if_else(.pred_female > threshold, "female", "male")) |>
    summarize(
      threshold = threshold,
      TP = sum(pred == "female" & sex == "female"),
      FP = sum(pred == "female" & sex == "male"),
      TN = sum(pred == "male"   & sex == "male"),
      FN = sum(pred == "male"   & sex == "female")
    )
}

thresholds <- c(0.2, 0.5, 0.8)
confusion_table <- map_dfr(thresholds, ~ confusion_at(penguins, .x))
confusion_table
```

Read as three separate 2x2 tables, threshold 0.5 looks like:

|              | Actual: female | Actual: male |
|--------------|----------------|--------------|
| Pred: female | TP             | FP           |
| Pred: male   | FN             | TN           |

## Performance Metrics

```{r metrics}
metrics_table <- confusion_table |>
  mutate(
    accuracy  = round((TP + TN) / (TP + FP + TN + FN), 3),
    precision = round(TP / (TP + FP), 3),
    recall    = round(TP / (TP + FN), 3),
    f1        = round(2 * precision * recall / (precision + recall), 3)
  ) |>
  select(threshold, TP, FP, TN, FN, accuracy, precision, recall, f1)

metrics_table
```

All three thresholds beat the 41.9% null error rate by a wide margin
(accuracy ranges roughly 91%–95%). Raising the threshold from 0.2 to 0.8
trades a couple of true positives for fewer false positives: precision rises
while recall drops slightly, which is exactly the tradeoff a threshold
controls.

## Threshold Use Cases

**A 0.2 threshold is preferable** when missing a true positive is far more
costly than a false alarm — e.g., a medical screening test flagging a
disease. You'd rather send a healthy patient for an unnecessary follow-up
(false positive) than send a sick patient home undiagnosed (false negative),
so the threshold for "flag as positive" should be low.

**A 0.8 threshold is preferable** when a false positive is the costly
mistake — e.g., an automated system that permanently bans a user account for
fraud. Acting on a wrong "positive" call is expensive and hard to undo, so
the model should only act when it's highly confident.

## Conclusions

The classifier comfortably beats the null baseline at every threshold tested,
confirming it has real predictive signal rather than just exploiting class
imbalance. Threshold choice is a business decision, not a modeling one: 0.2
maximizes recall (catches the most true positives, at the cost of more false
alarms) and 0.8 maximizes precision (fewer false alarms, at the cost of
missing a few more true positives). To extend this work: plot a full ROC
curve across all thresholds rather than just three points, and compute AUC as
a threshold-independent summary of model quality.

## AI Use

Anthropic. (2026). Claude (model: claude-sonnet-5) [Large language model].
[https://claude.ai/](https://claude.ai/) — used to write the confusion-matrix
and metrics computation code, verify the resulting numbers against a
Python reference implementation, and draft this write-up. Reviewed and
adapted by the author. `penguin_predictions.csv` itself came from the course
GitHub repository (not AI-generated).
